The host galaxy of the short GRB~111117A at z = 2.211\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}

### 2017-05-28 GRB111117A.tex
Implement Jures comments
Closes #1

\title{The host galaxy of the short GRB111117A at z = 2.211.
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 088.A-0051 and 091.D-0904.}}

\titlerunning{GRB111117A}

@@

\abstract{
Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to localize. We here present evidence of the, to date, highest redshift high-confidence short GRB, located at $z=2.211$. This value supersedes a previously estimated, lower redshift value. Spectroscopic observations along with a wealth of imaging data allow us to place tight constraints on the nature of the host galaxy at an unprecedented distance. 
The rest frame X-ray derived hydrogen column density is high compared to a complete sample of sGRBs and seems to follow the evolution with redshift as traced by the hosts of long GRBs (lGRBs). This could indicate that the part of the sGRB population hosted in late-type galaxies, live in similar environments to that of lGRB hosts. 
The host is in the brighter end of the expected host brightness distribution at $z = 2.211$, but only 30 per cent of the distribution should be missed due to host faintness at this redshift. This indicates that we are not missing redshifts for a dominant fraction of the sGRB hosts at this redshift. The redshift of GRB111117A is evidence against a lognormal delay-time model for sGRBs through the predicted redshift distribution of sGRBs, which is very sensitive to high-z sGRBs.
From the age of the universe at explosion time, an initial progenitor separation of a$_0 < 3.2 $R$_\odot$ is required for the case of a binary neutron star (NS) system. This puts constraints on the progenitor system evolution up to the time of explosion.

}

@@

There is now mounting evidence that most short-duration gamma-ray bursts (GRBs) come from the merger of neutron stars (NSs), either with another NS, or a black hole, due to their apparent association with kilonovae 
\citep{Barnes2013a, Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The
absence of associated supernovae in deep searches \citep[e.g.][]
{Hjorth2005a,Fox2005,Hjorth2005b} supports this idea and distinguishes the physical origin of sGRBs from their long-duration counterparts.
%\citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}. 
The range of host galaxy types, in some cases apparently absent, and typically more massive 
and on average less actively star-forming than long GRB hosts  \citep{Fong2013b}, as well as their positions within
hosts \citep{Fong2013a}, suggest an origin for the progenitors that can be very 
long-lived, and associated with stellar mass rather than purely 
star-formation rate \citep{Berger2014}.

This progress on the progenitors of sGRBs has been enabled by
localization by the \emph{Swift} satellite \citep{Gehrels2009a}. The bulk of these localizations have associated galaxies at relatively low redshifts with a median redshift $z\sim0.5$ \citep{Berger2014}. Because the redshifts are measured from the hosts, measurements of the redshift distribution of sGRBs is biased towards low redshift due to cosmological dimming. 
%at least partly due to the faintness of the bursts and their afterglows, though this is not clear. 
So far, the electromagnetic signal from sGRBs are the only means to accurately localize NS mergers and holds the promise for detection of an associated gravitational wave signal \citep{Ghirlanda2016}. From the burst and following afterglow, modeling the properties and mapping the environments additionally allows insights into the phenomenon itself.

@@

In this \emph{Letter} we present the spectrum of the host galaxy of the short
GRB\,111117A, showing it to be at higher redshift than previously
estimated. We present the rest frame burst properties based on this new distance
compared to previous analyses \citep{Margutti2012,Sakamoto2013} and revisit the host properties derived from the new solution to the SED fit. 
%While no optical afterglow was detected, the excellent localization from a detection
%of the X-ray afterglow by the \emph{Chandra X-ray Observatory} also allows
%us to discuss the positioning and environmental properties of this remarkably
%distant short GRB.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{Imaging of the field with the X-shooter slit overlaid. Only one slit is shown, despite 4 epochs of spectroscopic observations because of the similarity in position angle. The image is the FORS2 R-band image, for which the photometry is shown in Fig. \ref{fig:SED}. The blue asterisk indicates the GRB position as derived from the \emph{Chandra} observations in \citet{Sakamoto2013}}
	\label{fig:spec_setup}
\end{figure}

%(21000 \AA when the K-band blocking filter is used)
Spectroscopic observations were carried out using the cross-dispersed echelle spectrograph, VLT/X-shooter \citep{Vernet2011}, in four seperate epochs. The burst was followed up 38 hours after the Burst Alert Telescope (BAT) trigger under ESO programme 088.A-0051 (PI: Fynbo) and again later using a different ESO program 091.D-0904 (PI: Hjorth). X-shooter can cover the wavelength range from 3000 \AA~to 24800 \AA~across three spectroscopic arms, simultaneously illuminated through the use of dichroics. The bias-correction, flat-fielding, order tracing, wavelength calibration, rectification, and flux calibration is then carried out using the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} and managed by Reflex \citep{Freudling2013}. Because the spectra are curved across each detector, a rectification algorithm is employed which introduces correlations between neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the degree of correlation while conserving the maximal resolution. The observations are combined and extracted using scripts described in Selsing et al. 2017 (in prep.) and available online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The signal-to-noise of the continuum is the near-infrared arm is too low to use the optimal extraction algorithm \citep{Horne1986}, why an aperture containing the entire object has been chosen for which the flux is summed and the corresponding error is summed in quadrature.
An overview of the spectroscopic observations is given in Tab. \ref{tab:spec_overview}, and the position of the slit on the target is shown in Fig. \ref{fig:spec_setup}.

@@

We determine a redshift of $z = 2.211$ from from the simultaneous detection of emission lines interpreted as \lya, \oii, \hb, \oiii$\lambda$5007, and \ha, with \hb~detected at low significance ($\sim 3 \sigma$). \ha is only visible in the first epoch, due to the K-band blocking filter used for the remainder observations. The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and shows significant velocity structure along the slit. A drop in the continuum to the blue of the \lya~line further supports the inferred redshift.

@@

The multi-color spectral energy distribution (SED) is then fit by \citet{Bruzual2003} stellar population synthesis models based on a \citet{Chabrier2003} initial mass function in \emph{LePhare} \citep{Ilbert2006}, where the redshift if fixed to the spectroscopic value of $z=2.211$. The best model is obtained with an unreddened galaxy template, and returns physical parameters of luminosity ($M_B=-22.0\pm0.1$\,mag), stellar mass ($\log(M_{\star}/M_\odot) = 9.9\pm0.2$), single stellar population age ($\tau = 0.5_{-0.25}^{+0.5}$ Gyr) and star-formation rate ($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). %We show the SED fit in Fig. \ref{fig:SED}.

@@

In fact, the large $i-z$ color is mistakenly interpreted as a 4000\,\AA\,break driving the galaxy photometric redshift of both of these earlier works. It is this exactly this overestimate in $z$-band brightness that led the authors to erronously low estimates in their photo-$z$ for GRB~111117A. Using the revised photometry from Table~\ref{tab:phot_overview}, the photo-$z$ of the galaxy is $z_{\mathrm{phot}}=2.04_{-0.21}^{+0.19}$, consistent with the spectroscopic value at the 1~$\sigma$ confidence level.

@@

As already pointed out \citep{Margutti2012, Sakamoto2013}, GRB111117A very securely belongs the the short class of GRBs. Because the observed classification indicators, $T_{90}$ and hardness ratio, do not depend strongly on redshift \citep{Littlejohns2013a}, the updated redshift does not change this designation significantly. The intrinsic luminosity goes up, as visible in the light curve in Fig. \ref{fig:sxray_lightcurve}, but it still is sub-luminous compared to majority of long GRBs. 
\citet{Bromberg2013} investigated the degree to which the long and short population distributions overlap and quantified the certainty in the class membership. GRB111117A has $96_{-5}^{+3}$ percent probability of being short. This is compared to the other two highest-z short candidates, GRB060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7 \lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB090426 \citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$ per cent).

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/XLC_111117A_rest.pdf}
	\caption{Restframe XRT lightcurve of GRB111117A, compared to the general population of XRT lightcurves of GRBs. The grey shared region is a compilation of long GRB lightcurves where the color represents density and the light blue is other short GRB lightcurves for which redshifts has been determined. Despite the remarkably high redshift, the burst’s luminosity is comparable to the bulk of the short burst population and still subluminous compared to the lGRB population.}
	\label{fig:sxray_lightcurve}
\end{figure}

@@

We plot the recalculated $N_H$ in Fig. \ref{fig:NH_z} where we compare with the distributions of complete samples of both long and short GRBs. The long sample is from \citet{Arcodia2016} and the short sample is from \citet{DAvanzo2014a}. 17 of the 99 long bursts do not have redshifts and likewise for 5 out of 16 for the short sample. Bursts without redshifts have been excluded for both groups. GRB111117A occupies a unique position in Fig. \ref{fig:NH_z} with a the highest $N_H$ of all short burst. The short sample, excluding GRB111117A, is located at low redshifts(z < 1) and are found to populate a similar column density environment to long GRBs at similar redshifts \citep{DAvanzo2014a}. The inferred hydrogen column for GRB111117A seems to follow the trend with increasing $N_H$ as a function of redshift as found for the long GRB hosts \citep{Arcodia2016}.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_z.pdf}
	\caption{Rest frame, X-ray derived hydrogen column densities for GRB111117A compared to complete samples of both long and short populations. In red is shown the long sample of \citet{Arcodia2016}. The detections are replaced with contours for clarity and the limits are shown with arrows. In blue is the sample from \citet{DAvanzo2014a}. GRB111117A is an outlier from the short sample, but seems to follow the distribution of long GRBs.}
	\label{fig:NH_z}
\end{figure}

@@

From the clear host association, GRB111117A does not belong to the hostless class of GRBs \citep{Berger2010a} and because the host exhibits clear emission lines indicative of a population of relatively young stars, as the majority of short GRBs, the host is a late-type galaxy \citep{Fong2013b}.
The host of GRB111117A is entirely consistent in terms of stellar mass and stellar age with the general population of short GRB hosts($\left\langle M _* \right\rangle = 10^{10.1} M_{\odot}$ and $\left\langle \tau _* \right\rangle = 0.3 $ Gyr) \citet{Leibler2010}. The SFR, on the other hand, is $\sim$1 order of magnitude higher than the typical SFR for short GRB hosts galaxies \citep{Berger2014} and more similar to the SFRs found in the hosts of long GRBs at a corresponding redshift \citep{Kruhler2015}. Only two hosts in the sample of short GRBs compiled in \citet{Berger2014} have a more vigorous star formation, meaning that it is in the very upper end of the star formation distribution, while not a complete outlier. This is likely a selection effect, because a less star-forming galaxy would exhibit weaker emission lines, thus making the redshift harder to determine. Being a late-type host, both the stellar mass and sSFR are entirely consistent with the expected host of sGRBs\citep{Behroozi2014}.
That we detect \lya~is consistent with the SED-inferred absence of dust, despite the moderate stellar age and the high X-ray derived hydrogen column density which would suggest the opposite. The centroid of the Ly$\alpha$ emission is found to be redshifted by $\sim$ 220 km s$^{-1}$ with respect to systemic, which is similar to what is found for long GRB hosts \citep{Milvang-Jensen2012a} where the outflow is attributed to star formation.

@@

A single sGRB at high redshift does little in terms of constraints on the redshift distribution and therethrough the progenitor channels, but other short hosts could be missed because they are intrinsically fainter and thus this high-z event is only detected due to the brightness of the host. \citet{Berger2014} compiled a sample of sGRB host luminosities, normalized by the characteristic galaxy luminosity at their respective redshift, $L_B/L^{\star}_{B}$. 26 out of 39 hosts (66 per cent) in the sample has redshifts. To convert the SED-inferred $M_B$ of GRB111117A to $L_B/L^{\star}_{B}$, we use the characteristic absolute B-band magnitude of the Schechter function for the blue galaxies ($U - V < 0.25$) in the redshift window $2.0 \leq z \leq 2.5$ from \citet{Marchesini2007} and find $L_B/L^{\star}_{B} = 1.2$, which is brighter than 70 per cent of the hosts in \citet{Berger2014} with measured $L_B/L^{\star}_{B}$. If we assume that we are able to get emission-line redshifts from hosts with $R < 25 $mags \citep{Kruhler2012}, then we would have missed around 30\% (8 out of 26 from the sample of \citealt{Berger2014} with measured $L_B/L^{\star}_{B}$), if they were at the redshift of GRB111117A. If we assume that the hosts missing redshift in \citet{Berger2014} follow the redshift distribution of the hosts \textit{with} redshift, this implies that we are \textit{not} missing a dominant fraction of sGRB redshift at $z \approx 2$, due to host faintness.

The theoretical redshift distributions of sGRBs depends on the type of delay-time function used to model the progenitor system. The redshift of GRB111117A puts constraints on the type of delay-time models suitable for modeling. The likelihood preferred lognormal time delay models investigated in \citet{Wanderman2015} predicts a rate of sGRBs at $z = 2.211$, $\sim$ 2 orders of magnitude lower compared to peak rate (z = 0.9). It is stated in \citet{Wanderman2015} that this preference depends critically on the absence of non-collapsar sGRB at $z \gtrsim 1.2$. The redshift of the burst, on the other hand, is close to the expected peak in sGRB rate calculated for the power law models \citep{Behroozi2014, Wanderman2015}.

@@

At $z = 2.211$, the age of the universe is 2.97 Gyr. If the progenitor systems of sGRBs are the merger of two neutron stars, this sets a hard upper limit to the coalescence timescale for such a system. In the absence of other mechanisms, the timescale of the orbital decay of the system is set by the energy loss due to gravitational waves, which in turn is set by the mass of constituent compact objects and the separation of the two \citep{Postnov2014}. If we assume that the formation timescale of the first galaxies is short compared to the time since the big bang \citep{Richard2011} and if we assume a mass of 1.4 M$_\odot$ for each of the neutron stars at the time of system formation, this places a limit on the initial separation of the two neutronstars of a$_0 < 3.2 $R$_\odot$, where a$_0$ is the initial separation. Because we have the age of the stellar population, by using this as a limit to the orbital decay time we can put tighter constraints, a$_0 < 2.1 $R$_\odot$, but this does not account for the possibility there could be an underlying stellar population of older stars from a previous star-formation epoch, which is why we consider a$_0 < 3.2 $R$_\odot$ as a hard upper limit. The delay time between formation and explosion is well accomodated by the models of \citet{Belczynski2006}, although the longest formation channels are excluded. This is especially true given the late type nature of the host \citep{OShaughnessy2008}.

@@

We have in this letter provided a revised, spectroscopic redshift for GRB111117A based on emission lines discovered using an improved reduction scheme, setting it at $z = 2.211$. This value supersedes the previous photometric redshift of $z \sim 1.3$ \citep{Margutti2012, Sakamoto2013}. Part of the reason for the erroneous redshift estimate of previous authors is attributed to a discrepancy in the measured $z$-band magnitude.

The rest-frame parameters of the burst and the conditions of the burst environment have been recalculated using the new distance. The X-ray derived hydrogen column density towards GRB111117A is the highest within a complete sample of short hosts, but seems to trace the evolution with redshift as found for the hosts of long GRBs.

@@

Although a single burst carries little leverage in terms of constraining the redshift distribution of sGRB, the high redshift of GRB111117A needs to be accommodated. A lognormal delay time model predicts a very low volumetric density of bursts at $z = 2.211$, whereas a power law delay time model peaks near GRB111117A. If more sGRBs are at this redshift, but missed due to faintness of their hosts, a lognormal delay time model will be disfavored. Compared to a sample of short hosts, GRB111117A is more luminous than 70 per cent of the sample with measured luminosities. Of the host sample, for 30 per cent, we would be unable to determine a redshift, should they be at the redshift of GRB111117A. This implies that we are \textit{not} missing a dominant fraction of the sGRBs at $z \sim 2$.

@@

\begin{acknowledgements}
TK acknowledges support through the Sofja Kovalevskaja Award to P. Schady.
\end{acknowledgements}

>>>>

\title{The host galaxy of the short GRB~111117A at z = 2.211\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}}

\titlerunning{GRB~111117A}

@@

\abstract{
Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to localize. We here present a spectroscopic observation of the host galaxy of GRB 111117A and measure its redshift to be $z = 2.211$, making it the most distant high-confidence short GRB with a secure redshift measurement. This value supersedes a previously estimated, lower redshift value. Spectroscopic observations along with a wealth of imaging data allow us to place tight constraints on the nature of the host galaxy at an unprecedented distance. 
The rest frame X-ray derived hydrogen column density is high compared to a complete sample of sGRBs and seems to follow the evolution with redshift as traced by the hosts of long GRBs (lGRBs). This could indicate that the part of the sGRB population hosted in late-type galaxies live in similar environments to that of lGRB hosts. 
The host lies in the brighter end of the expected host brightness distribution at $z = 2.211$, and is actively star forming. Using the host as a benchmark for redshift determination, we assess that only 30 per cent of sGRB redshifts should be missed due to host faintness at this redshift. This indicates that we are not missing redshifts for a dominant fraction of the sGRB hosts at this redshift. The redshift of GRB~111117A is evidence against a lognormal delay-time model for sGRBs through the predicted redshift distribution of sGRBs, which is very sensitive to high-z sGRBs.
%From the age of the universe at explosion time, an initial progenitor separation of a$_0 < 3.2 $R$_\odot$ is required for the case of a binary neutron star (NS) system. This puts constraints on the progenitor system evolution up to the time of explosion.
}

@@

There is now mounting evidence that most short-duration gamma-ray bursts (GRBs) come from the merger of neutron stars (NSs), either with another NS, or a black hole, due to their apparent association with kilonovae \citep{Barnes2013a, Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The absence of associated supernovae in deep searches \citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b} supports this idea and distinguishes the physical origin of sGRBs from their long-duration counterparts.
%\citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}. 
The range of host galaxy types, in some cases apparently absent, and typically more massive 
and on average less actively star-forming than long GRB hosts  \citep{Fong2013b}, as well as their positions within hosts \citep{Fong2013a}, suggest an origin for the progenitors that can be very long-lived, and associated with stellar mass rather than purely star-formation rate \citep{Berger2014}. The bulk of sGRBs have associated galaxies at relatively low redshifts with a median $z\sim0.5$ \citep{Berger2014} and because most redshifts are measured from the hosts, this value is biased towards low $z$.
%This progress on the progenitors of sGRBs has been enabled by the afterglow detection and burst localization by the \emph{Swift} satellite \citep{Gehrels2009a}. The bulk of these localizations have associated galaxies at relatively low redshifts with a median $z\sim0.5$ \citep{Berger2014} and because most redshifts are measured from the hosts, this value is biased towards low $z$.

%So far, the electromagnetic signal from sGRBs are the only means to accurately localize NS mergers and holds the promise for detection of an associated gravitational wave signal \citep{Ghirlanda2016}. From the burst and following afterglow, modeling the properties and mapping the environments additionally allows insights into the phenomenon itself.

@@

In this \emph{Letter} we present the spectrum of the host galaxy of the short GRB 111117A at z=2.211, a significantly higher redshift than the previously estimated redshift based on photometric studies. We present the rest frame burst properties based on this new distance
compared to previous analyses \citep{Margutti2012,Sakamoto2013} and revisit the host properties derived from the new solution to the SED fit. 
%While no optical afterglow was detected, the excellent localization from a detection
%of the X-ray afterglow by the \emph{Chandra X-ray Observatory} also allows
%us to discuss the positioning and environmental properties of this remarkably
%distant short GRB.

@@

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/GRB111117A_spec_obs.pdf}
	\caption{Imaging of the field with the X-shooter slit overlaid. Only one slit is shown, despite 4 epochs of spectroscopic observations because of the similarity in position angle. The image is the FORS2 R-band image and%, for which the photometry is shown in Fig. \ref{fig:SED}. 
	the blue asterisk i ndicates the GRB position as derived from the \emph{Chandra} observations in \citet{Sakamoto2013}. The errorcircle of the \emph{Chandra}-position is too small to resolve in the image.}
	\label{fig:spec_setup}
\end{figure}

%(21000 \AA when the K-band blocking filter is used)
Spectroscopic observations were carried out using the cross-dispersed echelle spectrograph, VLT/X-shooter \citep{Vernet2011}, in four seperate epochs. The burst was followed up 38 hours after the Burst Alert Telescope (BAT) trigger under ESO programme 088.A-0051 (PI: Fynbo) and again later using a different ESO program 091.D-0904 (PI: Hjorth). X-shooter can cover the wavelength range from 3000 \AA~to 24800 \AA~across three spectroscopic arms, simultaneously illuminated through the use of dichroics. The bias-correction, flat-fielding, order tracing, wavelength calibration, rectification, and flux calibration is then carried out using the VLT/X-shooter pipeline version 2.8.4 \citep{Goldoni2006, Modigliani2010} and managed by Reflex \citep{Freudling2013}. Because the spectra are curved across each detector, a rectification algorithm is employed which introduces correlations between neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the degree of correlation while conserving the maximal resolution. The observations are combined and extracted using scripts described in Selsing et al. 2017 (in prep.) and available online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The signal-to-noise of the continuum in the near-infrared arm is too low to use the optimal extraction algorithm \citep{Horne1986} and therefore the extraction is carried out with a simple aperture.
An overview of the spectroscopic observations is given in Tab. \ref{tab:spec_overview}, and the position of the slit on the target is shown in Fig. \ref{fig:spec_setup}.

@@

We determine a redshift of $z = 2.211$ from the simultaneous detection of emission lines interpreted as \lya, \oii, \hb, \oiii$\lambda$5007, and \ha, with \hb~detected at low significance ($\sim 3 \sigma$). \ha~is only visible in the first epoch, due to the K-band blocking filter used for the remainder observations. The nebular lines exhibit a spatial extent of $\sim$ 1\farc5 and show significant velocity structure along the slit. A drop in the continuum to the blue of the \lya~line further supports the inferred redshift.

@@

The multi-color spectral energy distribution (SED) is then fit by \citet{Bruzual2003} stellar population synthesis models based on a \citet{Chabrier2003} initial mass function in \emph{LePhare} \citep{Ilbert2006}, where the redshift is fixed to the spectroscopic value of $z=2.211$. The best model is obtained with an unreddened galaxy template, and returns physical parameters of luminosity ($M_B=-22.0\pm0.1$\,mag), stellar mass ($\log(M_{\star}/M_\odot) = 9.9\pm0.2$), single stellar population age ($\tau = 0.5_{-0.25}^{+0.5}$ Gyr) and star-formation rate ($SFR_{\mathrm{SED}}=11_{-4}^{+9} M_\odot\,\mathrm{yr}^{-1}$). %We show the SED fit in Fig. \ref{fig:SED}.

@@

In fact, the large $i-z$ color is mistakenly interpreted as a 4000\,\AA\,break driving the galaxy photometric redshift of both of these earlier works. Using the revised photometry from Table~\ref{tab:phot_overview}, the photo-$z$ of the galaxy is $z_{\mathrm{phot}}=2.04_{-0.21}^{+0.19}$, consistent with the spectroscopic value at the 1~$\sigma$ confidence level.

@@

As already pointed out \citep{Margutti2012, Sakamoto2013}, GRB~111117A very securely belongs the short class of GRBs. Because the observed classification indicators, $T_{90}$ and hardness ratio, do not depend strongly on redshift \citep{Littlejohns2013a}, the updated redshift does not change this designation significantly. The intrinsic luminosity increases, as visible in the light curve in Fig. \ref{fig:sxray_lightcurve}, but it still is sub-luminous compared to majority of long GRBs. 
\citet{Bromberg2013} investigated the degree to which the long and short population distributions overlap and quantified the certainty in the class membership. GRB~111117A has $96_{-5}^{+3}$ percent probability of being short. Compared to the other two highest-z short candidates, GRB060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7 \lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB090426 \citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$ per cent), the certainty in class membership for GRB~111117A is much higher.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/XLC_111117A_rest.pdf}
	\caption{Restframe XRT lightcurve of GRB~111117A, compared to the general population of XRT lightcurves of GRBs. The grey shared region is a compilation of long GRB lightcurves where the color represents density and the light blue is other short GRB lightcurves for which redshifts has been determined. Despite the remarkably high redshift, the burst’s luminosity is comparable to the bulk of the short burst population and still subluminous compared to the lGRB population.}
	\label{fig:sxray_lightcurve}
\end{figure}

@@

We plot the recalculated $N_H$ in Fig. \ref{fig:NH_z} where we compare with the distributions of complete samples of both long and short GRBs. The long sample is from \citet{Arcodia2016} and the short sample is from \citet{DAvanzo2014a}. 17 of the 99 long bursts do not have redshifts and likewise for 5 out of 16 for the short sample. Bursts without redshifts have been excluded for both groups. GRB~111117A occupies a unique position in Fig. \ref{fig:NH_z} with the highest $N_H$ of all short burst. The short sample, excluding GRB~111117A, is located at low redshifts(z < 1) and are found to populate a similar column density environment to long GRBs at similar redshifts \citep{DAvanzo2014a}. The inferred hydrogen column for GRB~111117A seems to follow the trend with increasing $N_H$ as a function of redshift as found for the long GRB hosts \citep{Arcodia2016}.

\begin{figure}
	\centering
	\includegraphics[width=9cm]{figures/NH_z.pdf}
	\caption{Rest frame, X-ray derived hydrogen column densities for GRB~111117A compared to complete samples of both long and short populations. In red is shown the long sample of \citet{Arcodia2016}. The detections are replaced with contours for clarity and the limits are shown with arrows. In blue is the sample from \citet{DAvanzo2014a}. GRB~111117A is an outlier from the short sample, but seems to follow the distribution of long GRBs.}
	\label{fig:NH_z}
\end{figure}

@@

%From the clear host association, GRB~111117A does not belong to the hostless class of GRBs \citep{Berger2010a} and because the host exhibits clear emission lines indicative of a population of relatively young stars, 
As the majority of short GRBs \citep{Fong2013b}, the host of GRB~111117A is a late-type galaxy and is entirely consistent in terms of stellar mass and stellar age with the general population of short GRB hosts($\left\langle M _* \right\rangle = 10^{10.1} M_{\odot}$ and $\left\langle \tau _* \right\rangle = 0.3 $ Gyr) \citet{Leibler2010}. The SFR, on the other hand, is $\sim$1 order of magnitude higher than the typical SFR for short GRB hosts galaxies \citep{Berger2014} and more similar to the SFRs found in the hosts of long GRBs at a corresponding redshift \citep{Kruhler2015}. Only two hosts in the sample of short GRBs compiled in \citet{Berger2014} have a more vigorous star formation, meaning that it is in the very upper end of the star formation distribution, while not a complete outlier. This is likely a selection effect, because a less star-forming galaxy would exhibit weaker emission lines, thus making the redshift harder to determine. Being a late-type host, both the stellar mass and sSFR are entirely consistent with the expected host of sGRBs\citep{Behroozi2014}.
That we detect \lya~is consistent with the SED-inferred absence of dust, despite the moderate stellar age and the high X-ray derived hydrogen column density which would suggest the opposite. The centroid of the Ly$\alpha$ emission is found to be redshifted by $\sim$ 220 km s$^{-1}$ with respect to systemic, which is similar to what is found for long GRB hosts \citep{Milvang-Jensen2012a} where the outflow is attributed to star formation.

@@

A single sGRB at high redshift does little in terms of constraints on the redshift distribution and therethrough the progenitor channels, but other short hosts could be missed because they are intrinsically fainter and thus this high-z event is only detected due to the brightness of the host. \citet{Berger2014} compiled a sample of sGRB host luminosities, normalized by the characteristic galaxy luminosity at their respective redshift, $L_B/L^{\star}_{B}$. 26 out of 39 hosts (66 per cent) in the sample has redshifts. To convert the SED-inferred $M_B$ of GRB~111117A to $L_B/L^{\star}_{B}$, we use the characteristic absolute B-band magnitude of the Schechter function for the blue galaxies ($U - V < 0.25$) in the redshift window $2.0 \leq z \leq 2.5$ from \citet{Marchesini2007} and find $L_B/L^{\star}_{B} = 1.2$, which is brighter than 70 per cent of the hosts in \citet{Berger2014} with measured $L_B/L^{\star}_{B}$. If we assume that we are able to get emission-line redshifts from hosts with $R < 25$ mags \citep{Kruhler2012}, then we would have missed around 30\% (8 out of 26 from the sample of \citealt{Berger2014} with measured $L_B/L^{\star}_{B}$), if they were at the redshift of GRB~111117A. If we assume that the hosts missing redshift in \citet{Berger2014} follow the redshift distribution of the hosts \textit{with} redshift, this implies that we are \textit{not} missing a dominant fraction of sGRB redshift at $z \approx 2$, due to host faintness.

The theoretical redshift distribution of sGRBs depends on the type of delay-time function used to model the progenitor system. The redshift of GRB~111117A puts constraints on the type of delay-time models suitable for modeling. The likelihood preferred lognormal time delay models investigated in \citet{Wanderman2015} predicts a rate of sGRBs at $z = 2.211$, $\sim$ 2 orders of magnitude lower compared to peak rate (z = 0.9). It is stated in \citet{Wanderman2015} that this preference depends critically on the absence of non-collapsar sGRB at $z \gtrsim 1.2$. The redshift of the burst, on the other hand, is close to the expected peak in sGRB rate calculated for the power law models \citep{Behroozi2014, Wanderman2015}.

@@

At $z = 2.211$, the age of the universe is 2.97 Gyr. If the progenitor systems of sGRBs are the merger of two neutron stars, this sets a hard upper limit to the coalescence timescale for such a system. In the absence of other mechanisms, the timescale of the orbital decay of the system is set by the energy loss due to gravitational waves, which in turn is set by the mass of constituent compact objects and the separation of the two \citep{Postnov2014}. If we assume that the formation timescale of the first galaxies is short compared to the time since the big bang \citep{Richard2011} and if we assume a mass of 1.4 M$_\odot$ for each of the neutron stars at the time of system formation, this places a limit on the initial separation of the two neutron stars of a$_0 < 3.2 $R$_\odot$, where a$_0$ is the initial separation. 
Assuming our measured age of the stellar population ($\tau = 0.5$ Gyr) as a limit for the orbital decay time we can put a tighter constraint of a$_0 < 2.1 $R$_\odot$. However, this does not account for the possibility there could be an underlying stellar population of older stars from a previous star-formation epoch, which is why we consider a$_0 < 3.2 $R$_\odot$ as a hard upper limit. The delay time between formation and explosion is well accomodated by the models of \citet{Belczynski2006}, although the longest formation channels are excluded. This is especially true given the late type nature of the host \citep{OShaughnessy2008}.

@@

We have in this letter provided a revised, spectroscopic redshift for GRB~111117A based on emission lines discovered using an improved reduction scheme, setting it at $z = 2.211$. This value supersedes the previous photometric redshift of $z \sim 1.3$ \citep{Margutti2012, Sakamoto2013}. Part of the reason for the erroneous redshift estimate of previous authors is attributed to a discrepancy in the measured $z$-band magnitude.

The rest-frame parameters of the burst and the conditions of the burst environment have been recalculated using the new distance. The X-ray derived hydrogen column density towards GRB~111117A is the highest within a complete sample of short hosts, but seems to trace the evolution with redshift as found for the hosts of long GRBs.

@@

Although a single burst carries little leverage in terms of constraining the redshift distribution of sGRB, the high redshift of GRB~111117A needs to be accommodated. A lognormal delay time model predicts a very low volumetric density of bursts at $z = 2.211$, whereas a power law delay time model peaks near GRB~111117A. If more sGRBs are at this redshift, but missed due to faintness of their hosts, a lognormal delay time model will be disfavored. Compared to a sample of short hosts, GRB~111117A is more luminous than 70 per cent of the sample with measured luminosities. Of the host sample, for 30 per cent, we would be unable to determine a redshift, should they be at the redshift of GRB~111117A. This implies that we are \textit{not} missing a dominant fraction of the sGRBs at $z \sim 2$.

@@

\begin{acknowledgements}
TK acknowledges support through the Sofja Kovalevskaja Award to P. Schady. CT acknowledges support from a Spanish National Research Grant of Excellence under project AYA 2014-58381-P and funding associated to a Ramón y Cajál fellowship under grant number RyC-2012-09984.
\end{acknowledgements}
